---
title: "Bayesian Linear Regression"
subtitle: NYC Airbnb Price Prediction | DS4420 Final Project
author: Meixu Chen, Yuetong Yin
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
set.seed(42)
library(MASS)
library(tidyverse)
library(brms)

```

## Overview

This notebook implements Bayesian linear regression from scratch in R.
We use the **analytical posterior** — no sampling packages (no rstan, brms, etc.).

The key difference from OLS: instead of a single point estimate for beta,
we get a full posterior distribution, which lets us report credible intervals
that capture uncertainty in our price predictions.

---

## 1. Load and clean data

```{r load-data}
## 1. Load prepared modeling data

df <- read_csv("bayesian_model_data.csv", show_col_types = FALSE)
cat("Rows:", nrow(df), "\n")
print(names(df))

# Use the prepared split exported from Python
train <- df |> filter(split == "train")
val   <- df |> filter(split == "validation")
test  <- df |> filter(split == "test")

cat(sprintf("Train: %d | Val: %d | Test: %d\n", nrow(train), nrow(val), nrow(test)))

```


```{r }

# Keep the same variables used in the Python linear baseline
keep_cols <- c(
  "log_price", "price",
  "accommodates", "bedrooms", "bathrooms",
  "room_type", "property_type", "minimum_nights",
  "number_of_reviews", "review_scores_rating", "host_is_superhost"
)

train <- train |> dplyr::select(all_of(keep_cols))
val   <- val   |> dplyr::select(all_of(keep_cols))
test  <- test  |> dplyr::select(all_of(keep_cols))

# Convert categorical variables
train <- train |>
  mutate(
    room_type = factor(room_type),
    property_type = factor(property_type),
    host_is_superhost = factor(as.character(host_is_superhost))
  )

val <- val |>
  mutate(
    room_type = factor(room_type, levels = levels(train$room_type)),
    property_type = factor(property_type, levels = levels(train$property_type)),
    host_is_superhost = factor(as.character(host_is_superhost),
                               levels = levels(train$host_is_superhost))
  )

test <- test |>
  mutate(
    room_type = factor(room_type, levels = levels(train$room_type)),
    property_type = factor(property_type, levels = levels(train$property_type)),
    host_is_superhost = factor(as.character(host_is_superhost),
                               levels = levels(train$host_is_superhost))
  )

# Standardize numeric variables using training statistics only
NUM_FEATS <- c(
  "accommodates", "bedrooms", "bathrooms",
  "minimum_nights", "number_of_reviews", "review_scores_rating"
)

train_means <- sapply(train[, NUM_FEATS], mean)
train_sds   <- sapply(train[, NUM_FEATS], sd)
train_sds[train_sds == 0] <- 1

for (col in NUM_FEATS) {
  train[[col]] <- (train[[col]] - train_means[col]) / train_sds[col]
  val[[col]]   <- (val[[col]]   - train_means[col]) / train_sds[col]
  test[[col]]  <- (test[[col]]  - train_means[col]) / train_sds[col]
}
```

## 3. Build design matrix

```{r design-matrix}
# Same feature structure as the Python linear baseline
feature_formula <- log_price ~ accommodates + bedrooms + bathrooms +
  room_type + property_type + minimum_nights +
  number_of_reviews + review_scores_rating + host_is_superhost

X_train <- model.matrix(feature_formula, data = train)
X_val   <- model.matrix(feature_formula, data = val)
X_test  <- model.matrix(feature_formula, data = test)

y_train <- train$log_price
y_val   <- val$log_price
y_test  <- test$log_price

cat("Design matrix:\n")
cat(sprintf("  Train: %d x %d\n", nrow(X_train), ncol(X_train)))
cat(sprintf("  Val:   %d x %d\n", nrow(X_val), ncol(X_val)))
cat(sprintf("  Test:  %d x %d\n", nrow(X_test), ncol(X_test)))
```

## 4. Bayesian linear regression — analytical posterior

For Bayesian linear regression with a normal prior on beta,
the posterior has a closed-form solution. This is exact, not approximate.

**Prior:** beta ~ N(0, tau^2 * I), where tau = 10

**Posterior mean and covariance:**
```{r bayesian-posterior}
tau2 <- 100
p    <- ncol(X_train)

# OLS estimate for sigma initialization
beta_ols  <- as.vector(ginv(t(X_train) %*% X_train) %*% t(X_train) %*% y_train)
sigma_est <- sd(y_train - X_train %*% beta_ols)
sigma2    <- sigma_est^2

cat(sprintf("Estimated sigma: %.4f\n", sigma_est))

A          <- t(X_train) %*% X_train + (sigma2 / tau2) * diag(p)
A_inv      <- ginv(A)
mu_post    <- as.vector(A_inv %*% t(X_train) %*% y_train)
Sigma_post <- sigma2 * A_inv

cat("Posterior computed successfully.\n")

```

## 5. Posterior coefficient summary

```{r coef-summary}
# 95% credible interval for each coefficient: mean +/- 1.96 * sd
post_sd <- sqrt(diag(Sigma_post))

coef_summary <- data.frame(
  feature = colnames(X_train),
  mean    = mu_post,
  sd      = post_sd,
  lower95 = mu_post - 1.96 * post_sd,
  upper95 = mu_post + 1.96 * post_sd
) |> arrange(desc(abs(mean)))

coef_summary[, -1] <- round(coef_summary[, -1], 4)
print(coef_summary)
```

## 6. Draw samples from posterior and make predictions

```{r posterior-predict}
n_draws <- 2000

# Draw beta samples from posterior
beta_samples <- mvrnorm(n = n_draws, mu = mu_post, Sigma = Sigma_post)

# Predictions on validation set
pred_mat_val <- X_val %*% t(beta_samples)
pred_mat_val <- pred_mat_val + matrix(
  rnorm(nrow(X_val) * n_draws, 0, sigma_est),
  nrow = nrow(X_val)
)

val_pred_mean <- rowMeans(pred_mat_val)
val_ci_lower  <- apply(pred_mat_val, 1, quantile, 0.025)
val_ci_upper  <- apply(pred_mat_val, 1, quantile, 0.975)

# Predictions on test set
pred_mat_test <- X_test %*% t(beta_samples)
pred_mat_test <- pred_mat_test + matrix(
  rnorm(nrow(X_test) * n_draws, 0, sigma_est),
  nrow = nrow(X_test)
)

test_pred_mean <- rowMeans(pred_mat_test)
test_ci_lower  <- apply(pred_mat_test, 1, quantile, 0.025)
test_ci_upper  <- apply(pred_mat_test, 1, quantile, 0.975)

cat("Predictions done.\n")
```

## 7. Evaluation metrics

```{r evaluation}
calc_metrics <- function(actual_log, pred_log, lower = NULL, upper = NULL) {
  resid <- actual_log - pred_log
  
  mae  <- mean(abs(resid))
  rmse <- sqrt(mean(resid^2))
  r2   <- 1 - sum(resid^2) / sum((actual_log - mean(actual_log))^2)
  
  actual_price <- exp(actual_log)
  pred_price   <- exp(pred_log)
  
  mae_price  <- mean(abs(actual_price - pred_price))
  rmse_price <- sqrt(mean((actual_price - pred_price)^2))
  
  out <- tibble(
    MAE_log = mae,
    RMSE_log = rmse,
    R2 = r2,
    MAE_price = mae_price,
    RMSE_price = rmse_price
  )
  
  if (!is.null(lower) && !is.null(upper)) {
    coverage  <- mean(actual_log >= lower & actual_log <= upper)
    avg_width <- mean(upper - lower)
    out <- out |>
      mutate(CI_coverage = coverage, Avg_CI_width = avg_width)
  }
  
  out
}

val_metrics  <- calc_metrics(y_val,  val_pred_mean,  val_ci_lower,  val_ci_upper)
test_metrics <- calc_metrics(y_test, test_pred_mean, test_ci_lower, test_ci_upper)

val_metrics
test_metrics
```

## 8. Credible interval plot

```{r ci-plot, fig.height=4, fig.width=10}
set.seed(1)

plot_df_ci <- tibble(
  y_true   = y_test,
  y_pred   = test_pred_mean,
  ci_lower = test_ci_lower,
  ci_upper = test_ci_upper
)

if (nrow(plot_df_ci) > 100) {
  plot_df_ci <- plot_df_ci |> slice_sample(n = 100)
}

plot_df_ci <- plot_df_ci |>
  arrange(y_true) |>
  mutate(idx = row_number())

ggplot(plot_df_ci, aes(x = idx)) +
  geom_linerange(
    aes(ymin = ci_lower, ymax = ci_upper),
    color = "steelblue", alpha = 0.5, linewidth = 0.7
  ) +
  geom_point(aes(y = y_pred), color = "steelblue", size = 1.5) +
  geom_point(aes(y = y_true), color = "black", size = 1, shape = 4) +
  labs(
    title = "Bayesian Predictions with 95% Credible Intervals",
    subtitle = "Blue = posterior mean + CI | X = true log_price",
    x = "Observation (sorted by true price)",
    y = "log_price"
  ) +
  theme_minimal()
```

```{r}

plot_df_log <- tibble(
  actual = y_test,
  predicted = test_pred_mean
)

ggplot(plot_df_log, aes(x = actual, y = predicted)) +
  geom_point(alpha = 0.2, size = 1.2, color = "steelblue") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Bayesian Regression: Actual vs Predicted (Log Scale)",
    x = "Actual log_price",
    y = "Predicted log_price"
  ) +
  theme_minimal()


plot_df_price <- tibble(
  actual = exp(y_test),
  predicted = exp(test_pred_mean)
)

ggplot(plot_df_price, aes(x = actual, y = predicted)) +
  geom_point(alpha = 0.2, size = 1.2, color = "steelblue") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Bayesian Regression: Actual vs Predicted (Price Scale)",
    x = "Actual Price",
    y = "Predicted Price"
  ) +
  theme_minimal()



```


```{r}

ggplot(tibble(residual = y_test - test_pred_mean), aes(x = residual)) +
  geom_histogram(bins = 60, fill = "coral", color = NA) +
  labs(
    title = "Bayesian Regression: Residual Distribution",
    x = "Residual",
    y = "Count"
  ) +
  theme_minimal()

set.seed(1)

plot_df_ci <- tibble(
  y_true   = y_test,
  y_pred   = test_pred_mean,
  ci_lower = test_ci_lower,
  ci_upper = test_ci_upper
)

if (nrow(plot_df_ci) > 100) {
  plot_df_ci <- plot_df_ci |> slice_sample(n = 100)
}

plot_df_ci <- plot_df_ci |>
  arrange(y_true) |>
  mutate(idx = row_number())



```

## 9. Coefficient credible interval plot

```{r coef-plot, fig.height=9}
coef_plot <- coef_summary |>
  filter(feature != "bias") |>
  arrange(mean)

ggplot(coef_plot, aes(x = mean, y = reorder(feature, mean))) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  geom_errorbarh(aes(xmin = lower95, xmax = upper95),
                 height = 0.3, color = "steelblue", linewidth = 0.7) +
  geom_point(color = "steelblue", size = 2) +
  labs(
    title = "Posterior mean and 95% credible intervals for each coefficient",
    x     = "Coefficient value (log_price scale)",
    y     = NULL
  ) +
  theme_minimal()
```

## 10. Comparison table

Fill in OLS and NN results from the Python notebooks.

```{r comparison-table}
comparison <- tibble(
  Metric = c(
    "MAE (log)", "RMSE (log)", "R2",
    "MAE ($)", "RMSE ($)", "95% CI coverage", "Avg CI width"
  ),
  OLS = c(
    "0.3968", "0.4951", "0.5045",
    "$72.40", "$110.31", "-", "-"
  ),
  Neural_Network = c(
    "0.3009", "0.4054", "0.7261",
    "$71.94", "$155.88", "-", "-"
  ),
  Bayesian = c(
    round(test_metrics$MAE_log, 4),
    round(test_metrics$RMSE_log, 4),
    round(test_metrics$R2, 4),
    paste0("$", round(test_metrics$MAE_price, 2)),
    paste0("$", round(test_metrics$RMSE_price, 2)),
    round(test_metrics$CI_coverage, 3),
    round(test_metrics$Avg_CI_width, 4)
  )
)

knitr::kable(comparison, caption = "Model comparison — test set")
```
